Future Cosmic Microwave Background delensing with galaxy surveys.

### 2017-07-11 Notes/galaxy_delensing.tex
text: conclusions

For ongoing experiments, we find that combining LSS tracers with CMB internal reconstruction will be particularly beneficial.
An optical survey like DES is able to remove $14\%$ of the B-mode lensing power alone, and together with WISE($8\%$) and the CIB ($27\%$) will allow to remove $36\%$ of the power using only LSS surveys.
Depending on the CMB instrumental noise and the amount of galactic foreground cleaning, delensing using LSS survey can correspond to an improvement of 30\% in the constraint on the tensor to scalar ratio $r$.
In the future, the decreasing level of instrumental noise of CMB experiments will dramatically improve its internal reconstruction of the structures lensing the CMB. 
The fraction of removed lensing B-modes will rapidly improve from the current level of Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) level. 
Already for Stage-3, the CMB internal reconstruction will be the main source of delensing. However it will still be less efficient than galaxies at delensing small scales $\ell>500$. For this reason, combining galaxy survey with the CMB will push the fraction of removed power from $57\%$ to $68\%$ for 3G.
Even for CMB S4, galaxy surveys will still play an important role.
For example, a tomographically binned LSST-survey will remove 56\% of the lensing power. While this performance is comparable to a Stage-3 CMB it will allow to probe the robustness against systematics of an eventual detection.
Indeed delensing with just CMB data will require a careful study of possible biases and systematics effects due to the use of the same source (the CMB) that we are trying to delens to reconstruct the lensing effect itself \cite{carron:2017,sehgal:2017,namikawa:2017}.
For this reason, efficient galaxies tracers are not only useful in the short term but will also play a role in cross-checking these internal biases in the future.

>>>>

For ongoing experiments, we find that LSS tracers will be particularly beneficial.
An optical survey like DES is able to remove $14\%$ of the B-mode lensing power alone, and together with WISE($8\%$) and the CIB ($27\%$) will allow to remove $36\%$ of the power using only LSS surveys.
Depending on the CMB instrumental noise and the amount of galactic foreground cleaning, delensing using LSS survey can correspond to an improvement of 30\% in the constraint on the tensor to scalar ratio $r$ compared to the value before delensing.
In the future, the decreasing level of instrumental noise in CMB experiments will dramatically improve its internal reconstruction of the structures lensing the CMB. 
The fraction of removed lensing B-modes will rapidly improve from the current expected levels for Planck ($8\%$) and SPTPol ($35\%$) to 3G (56\%) and CMB S4 (85\%) levels. 
Already for Stage-3, the CMB internal reconstruction will be the main source of delensing. However, it will still be less efficient than galaxies at tracing the lensing potential at small scales $\ell>500$. For this reason, combining galaxy survey with the CMB will push the fraction of removed power from $57\%$ to $68\%$ for 3G.
Even for CMB S4, galaxy surveys will still play an important role.
For example, a tomographically binned LSST-like survey by itself will remove 56\% of the lensing power. This performance is lower than a S4 CMB internal reconstruction and comparable to a Stage-3 CMB. 
However it will allow us to probe the robustness against systematics of an eventual detection of primordial gravitational waves.
Indeed delensing with just CMB data will require a careful study of possible biases and systematics effects due to the use of the same source (the CMB) that we are trying to delens to reconstruct the lensing effect itself \cite{carron:2017,sehgal:2017,namikawa:2017}.
For this reason, efficient galaxies tracers are not only useful in the short term but, in the future, will also play a role in testing internal biases and performing consistency tests.
